---
title: "Week 6 Exercises"
output: html_document
---

# Exercise 1

The data frames df_w and df_f represent repeated measures data from 60 participants. Variables F1-F3 and W1-W3 are "sub-variables" that will be used to make two composite variables F_total and W_total, respectively. 

```{r include=F}
library(tidyverse)
library(broom)

# simulating data
set.seed(1)
df_f <- data.frame(id = rep(1:60, each = 2) %>% as.character,
                 session = rep(2:1, 60),
                 group = rep(1:2, each = 60),
                 F1 = sample.int(4, 120, replace = T) - 1,
                 F2 = sample.int(4, 120, replace = T) - 1,
                 F3 = (sample.int(4, 120, replace = T) - 1)) %>%
  mutate(F3 = F3*(session-1), F2 = F2*(group-1)*(2-session)) %>%
  mutate(group = factor(group))

df_w <- data.frame(id = rep(1:60, each = 2) %>% as.character,
                 session = rep(1:2, 60) %>% factor,
                 group = rep(1:2, each = 60) %>% factor,
                 W1 = sample.int(4, 120, replace = T) - 1,
                 W2 = sample.int(4, 120, replace = T) - 1,
                 W3 = sample.int(4, 120, replace = T) - 1)

str(df_f)
str(df_w)
```

#### 1.1

Merge the two data frames together.

```{r}
df_f <- df_f %>%
  mutate(session=factor(session))
df_fw <- full_join(df_f, df_w, by = c("id", "session", "group"))

```

#### 1.2

Using the merged data frame, create the **composite variables F_total and W_total, which are the sums of F1-F3 and W1-W3, respectively** (i.e. their values can range from 0 to 9).

```{r}

df_fw <- df_fw %>%
  mutate(F_total = rowSums(.[4:6])) %>%
  mutate(W_total = rowSums(.[7:9]))

```


# Exercise 2

#### 2.1

Visualise the distributions of F_total and W_total for the two groups and measurement sessions (for example as boxplots). 

```{r}

df_fw %>%
  ggplot(aes(F_total)) +
  geom_boxplot() +
  facet_wrap(group~session)

df_fw %>%
  ggplot(aes(W_total)) +
  geom_boxplot() +
  facet_wrap(group~session)

```

#### 2.2

Fit a linear regression model with F_total as the DV, and session and group as predictors.  

```{r}
str(df_fw)
F.model <- lm(F_total ~ group + session, data = df_fw)

summary(F.model)

```

#### 2.3

Look at the means of F_total by group and session. How are they linked to the linear regression model coefficients?

```{r}
df_fw <- df_fw %>%
  group_by(group, session) %>%
  mutate(F_mean = mean(F_total)) %>%
  ungroup()

df_fw %>%
  ggplot(aes(session, F_mean, color = group)) +
  geom_point()

# The mean of F_total goes down between sessions for one group, and up for the other. So it seems there is an interaction effect. We can see if it's significant by adding it to the regression model.

F.model.2 <- lm(F_total ~ group + session + group * session, data = df_fw)

summary(F.model.2)


```


# Exercise 3

Visualise the [anscombe dataset](https://rdrr.io/r/datasets/anscombe.html) using ggplot2.

```{r}
summary(anscombe)

x1y1 <- lm(data = anscombe, y1 ~ x1)

x1y1

ggplot(data = anscombe, aes(x1, y1)) +
  stat_summary(fun.data=mean_cl_normal) + 
  geom_smooth(method='lm', formula= y~x)

x2y2 <- lm(data = anscombe, y2 ~ x2)

x2y2

ggplot(data = anscombe, aes(x2, y2)) +
  stat_summary(fun.data=mean_cl_normal) + 
  geom_smooth(method='lm', formula= y~x)

x3y3 <- lm(data = anscombe, y3 ~ x3)

x3y3

ggplot(data = anscombe, aes(x3, y3)) +
  stat_summary(fun.data=mean_cl_normal) + 
  geom_smooth(method='lm', formula= y~x)

x4y4 <- lm(data = anscombe, y4 ~ x4)

x4y4

ggplot(data = anscombe, aes(x4, y4)) +
  stat_summary(fun.data=mean_cl_normal) + 
  geom_smooth(method='lm', formula= y~x)


```


# Exercise 4

Prepare a **separate R Notebook/Markdown document**, which will be the first draft of your final assignment with your own data. In the draft, include the following:  

1. Outline the study design, your research question, DV(s), IV(s)  
2. Data wrangling: start from reading in the raw data and show all steps  
3. Plot your main result(s)    

Even if you had already completed some of these steps before, please include all of them in your document. **NOTE: Return either a readable HTML document (.html or .nb.html), or an .Rmd file along with your data, to make it possible for us to review your work!** Make the document as professional-looking as possible (you can, of course, include your comments/questions in the draft). You will get feedback on the draft, based on which you can then make the final version. The final document should be a comprehensive report of your data wrangling process and results.  