---
title: "Week 2 Exercises"
author: Henrik Husberg
output:
  html_document:
    df_print: paged
    toc: true
---

# Exercise 1

#### 1.1

Create a new code chunk where you load the tidyverse package. In the chunk settings, suppress any output messages.  

```{r include=F}
library(tidyverse)
library(ggplot2)
```

#### 1.2

The tibble *df* has 60 observations (rows) of variables (columns) group, gender, age, score1 and score2 (continuous scores from two tests). Each row represents one participant.  
```{r include = F}
# this chunk generates the data, and you can ignore it
set.seed(1)
df  <- tibble(group = sample(c(rep(1:2, each=30))),
              gender = sample(c(rep(c('M', 'F'), 28), rep(999,4))),
              score1 = rnorm(60, 18, 2),
              score2 = as.character(runif(60, 0.7, 1.1)*score1))
  
```

```{r}
df
```


There is something to fix in three of the variables. Explore the data and describe what needs to be corrected.  

**Hint**: You can use e.g. str(), distinct(), and summary() to explore the data.  
```{r}
## Henrik: Here I'll take a look at the structure of the data to see if I can spot any oddities:
str(df)
summary(df)
distinct(df, group)
distinct(df, gender)
```
Henrik: Looking at the data, we see that the variable score2 seems to be coded as a string, not numerical. So we need to convert score2 to numerical data.
In the gender variable we have some apparently missing data coded as 999, we will convert those to NA:s. Also, the gender variable should be coded as a factor.


# Exercise 2

#### 2.1

Make the corrections you described above.
```{r}
##Alt: df <- transform(df, score2 = as.numeric(score2))
df <- df %>%
  mutate(score2 = as.numeric(score2)) %>%
  mutate(gender = na_if(gender, 999)) %>%
  mutate(gender = as.factor(gender)) 
  
summary(df)
str(df)
```
Henrik: Now the variable score2 is coded as numerical data, group as integer, gender as a factor with two leves and missing values as NA and we can use the data in our further analyses.

#### 2.2

Count observations by group and gender. Arrange by the number of observations (ascending).
```{r}
df %>%
  count(group, gender) %>%
  arrange(n)

```


# Exercise 3

#### 3.1

Create a new variable, **score_diff**, that contains the difference between score1 and score2.

```{r}
df <- df %>%
  mutate(score_diff = score2 - score1)
str(df)
```

#### 3.2

Compute the means of score1, score2, and score_diff.

**Hint**: Like mutate(), summarise() can take multiple variables in one go.

```{r}
df %>%
  summarise(mean1 = mean(score1), mean2 = mean(score2), mean_diff = mean(score_diff))

```

#### 3.3

Compute the means of score1, score2, and score_diff **by gender**.

```{r}
df %>%
  group_by(gender) %>%
  summarise(mean1 = mean(score1), mean2 = mean(score2), mean_diff = mean(score_diff)) %>%
  ungroup(df)

```


# Exercise 4

#### 4.1

Using ggplot2, create a scatter plot with score1 on the x-axis and score2 on the y-axis. 
```{r}
df %>%
  ggplot(aes(x=score1, y=score2)) + geom_point()
```


#### 4.2

Continuing with the previous plot, colour the points based on gender. 

Set the output figure width to 10 and height to 6.


```{r}
df %>%
  ggplot(aes(x=score1, y=score2, color=gender), width = 10, height = 6) + geom_point()
```


# Exercise 5

#### 5.1

Add the author (your name) and date into the metadata section. Create a table of contents.  

#### 5.2

Knit your document to HTML by changing `html_notebook` to `html_document` in the metadata, and pressing Knit.   
