---
title: "Week 3 Exercises"
output: html_document
---

# Exercise 1
```{r}
library(tidyverse)
library(ggplot2)
```
#### 1.1

Read the *region_scores.csv* data into R. It should have 240 observations of 6 variables.  

```{r}
getwd()
df <- readr::read_csv('region_scores.csv', locale = locale(encoding = "ISO-8859-1"))
df %>%
  distinct(region)
```


#### 1.2

Cut the continuous variable **age** into a categorical variable **age_group** that has groups of width 10. Use ggplot2's cutting function.  

Count observations by age group.

```{r}
df <-df %>%
  mutate(age_group = cut_width(age, 10))
df %>%
  count(age_group)
str(df)
```


# Exercise 2

The chunk below is supposed to produce a plot but it has some errors.  

The figure should be a scatter plot of **age** on the x-axis and **score** on the y-axis, with points coloured by **education level** (3 levels). It should also have three linear regression lines, one for each of the education levels.  

Fix the code to produce the right figure.  

What happens if you use `geom_jitter()` instead of `geom_point()`?  

**Hint**: Examine the code bit by bit: start by plotting just the scatter plot without `geom_smooth()`, and add the regression lines last.  

```{r}
str(df)
## The education variable is numerical, so first I convert it to factor:
df <- df %>%
  mutate(education = as_factor(education))
str(df)
## Now the data should be ok for the required plot
## For plot 1 I corrected the code so that the x and y-axes as well as the coloring are defined inside the aes-function
p1 <- df %>% 
  ggplot(aes(x=age, y=score, color=education)) + geom_point() + geom_smooth(method = "lm")
## For the second plot I change geom_point to geom_jitter
p2 <- df %>% 
  ggplot(aes(x=age, y=score, color=education)) + geom_jitter() + geom_smooth(method = "lm")
## Finally, I compare the plots
p1
p2
## In plot 2, the points are randomly spread out a bit on the y-axis so as to make it easier to spot the different points.
```


# Exercise 3

#### 3.1

Calculate the mean, standard deviation (sd), and number of observations (n) of **score by education and age group**. Also calculate the standard error of the mean (by using sd and n). Save these into a new data frame (or tibble) named edu_age_stats.  

```{r}
edu_age_stats <- df %>%
  group_by(education, age_group) %>%
  summarise(mean_score = mean(score),
            sd_score = sd(score),
            n = n()) %>%
  mutate(se_score = sd_score/sqrt(n)) %>%
  ungroup()
str(edu_age_stats)
```

#### 3.2

Using edu_age_stats, plot a bar plot that has **age group** on the x-axis, **mean score** on the y-axis, and **education levels** in subplots (facets).  

Use `geom_errorbar()` to add error bars that represent standard errors of the mean.  

```{r}
edu_age_stats %>%
  ggplot(aes(age_group, mean_score)) + geom_bar(stat='identity') + facet_wrap(~education) + geom_errorbar(aes(ymin=mean_score-se_score, ymax=mean_score+se_score))
```


# Exercise 4

#### 4.1

Create a figure that has boxplots of **score** (x-axis) by **region** (y-axis).  

```{r}
df %>%
  ggplot(aes(score, region)) + geom_boxplot()
```

#### 4.2

Group the data by region and add mean score by region to a new column **mean_score**. Do this with **`mutate()`** (not `summarise()`). 

Reorder the levels of region based on mean_score.  

**Hint**: Remember to ungroup after creating the mean_score variable.  

```{r}
df_grp <- df %>%
  group_by(region) %>%
  mutate(mean_score=mean(score)) %>%
  ungroup()
str(df_grp)
df_grp %>%
  mutate(region_reorder = fct_reorder(region, mean_score)) %>%
  ggplot(aes(score, region_reorder)) + geom_boxplot()
```


# Exercise 5

Using the data you modified in exercise 4.2, plot **mean scores** (x-axis) by **region** (y-axis) as points. The regions should be ordered by mean score.  

Use `stat_summary()` to add error bars that represent standard errors of the mean.  

**Hint**: Be careful which variable - mean_score or score - you're plotting in each of the geoms.  

```{r}
str(df_grp)
df_grp %>%
  mutate(region_reorder = fct_reorder(region, mean_score)) %>%
  ggplot(aes(mean_score, region_reorder)) + stat_summary(geom = "point", fun.data = "mean_se") + stat_summary(geom = "linerange", fun.data = "mean_se")
## Using **mean_score**, there is no variation as there is a single score for each region. Using **score** we would get error bars.

```


