---
title: "Week 5 Exercises"
output: html_notebook
---

Read the file timeuse_tidy.rds with `readRDS()`. The file contains the dataset that we tidied in the exercise session: records of daily time use from participants over multiple days. Note that since the data has been stored as rds (R-specific format), column types and factor levels are as we left them, and don't need to be re-corrected.

```{r}
library(cowplot)
library(tidyverse)
library(lubridate)
time_d <- readRDS("timeuse_tidy.rds")

```


# Exercise 1

#### 1.1

Create a new variable that contains combined activity classes: "Work or school" (Work, Business, Education), "Free time" (Shopping, Social / Leisure, Home, Vacation), and "Other".  

Calculate the mean time spent on each of the combined activity classes, grouped by **weekday, participant ID, and occ_full_time**.  

```{r}
time_d <- time_d %>%
  mutate(activity3 = case_when(activity_class %in% c("Work", "Business", "Education") ~ 1, activity_class %in% c("Shopping", "Social / Leisure", "Home", "Vacation") ~ 2, TRUE ~ 3)) %>%
  mutate(activity3 = factor(activity3))
str(time_d)
levels(time_d$activity3) <- c("Work or school", "Free time", "Other")
levels(time_d$activity3)

time_d_agg <- time_d %>%
  group_by(weekday, indivID, occ_full_time, activity3) %>%
  summarise(time_spent = mean(time_spent),
            n = n()) %>%
  ungroup()
time_d_agg
  
```

#### 1.2 

Visualise the means you calculated.

```{r}
time_d_agg %>%
  ggplot(aes(weekday, time_spent, color = activity3)) +
  geom_col() + 
  facet_wrap(~occ_full_time)

```


# Exercise 2

#### 2.1

What is computed in the code chunk below - what do the numbers tell you?  

Can you think of another way to calculate the same thing?  

```{r}

dist_date <- tidytime %>%
  distinct(indivID, date) %>%
  arrange(date) %>%
  count(date)

# Calculates the number of individuals for which there are observations on the dates. So for 2016-10-14 there are 6 observations from 6 individuals.

```

#### 2.2

Plot the numbers from above (use points, lines, or whatever you think is suitable).  

```{r}
dist_date <- tidytime %>%
  distinct(indivID, date) %>%
  arrange(date) %>%
  count(date)

dist_date %>%
  ggplot(aes(date, n)) + 
  geom_line() +
  scale_x_date(date_breaks = "1 week", date_labels = "%W", date_minor_breaks = "1 day")

```


# Exercise 3

#### 3.1

Count the **total number of participants** in the data.  

```{r}
tidytime %>%
  distinct(indivID) %>%
  count()
```

#### 3.2

For each participant, count the **number of separate days** that they recorded their time use on.  

```{r}
tidytime %>%
  group_by(indivID) %>%
  distinct(date) %>%
  count()
```


# Exercise 4

Explain step by step what happens in the code chunk below, and what the final figure represents.

```{r fig.height=20, fig.width=10}
tidytime %>% # Pipe the data to:
  group_by(indivID) %>% # Group by individual participant
  mutate(start_date = min(date)) %>% # Create new variable, start_date, by the smallest recorded value of date
  ungroup %>% # Remove grouping
  mutate(indivID = factor(indivID), 
         # make id variable a factor
         indivID = fct_reorder(indivID, start_date) %>%
           # Reorder participants according to start_date
           fct_rev()) %>% # Reverse order (why?)
  ggplot(aes(date, indivID, colour = month(start_date, label = T))) + 
  # Plot with date on x-axis, indivID on y-axis
  geom_line() + 
  # Plot lines
  geom_point(size=.5, alpha=.1) +
  # Plot points, larger than lines
  theme_bw() + 
  # Set theme
  scale_y_discrete(breaks = "none") +
  # Remove id's from y-axis I think
  labs(x = "Date", y = "", colour = "Starting month")
# Set label for x-axis to Date, none for y-axis, headline color-info with Starting month
# The final figure lines represents the time-span of observations for each participant, so we see when they started and finished. The points also illustrate missing data.

```





