---
title: "dataViz_lecture5"
author: "Ben Cowley"
date: "07/10/2020"
output: 
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '4'
---
# Data Viz Lecture 5

### Importing data
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# attach packages
library(tidyverse)
library(here)

# set working directory - can be convenient if reading multiple datasets from files, 
# to put them in a folder 'data' and uncomment the next line
# knitr::opts_knit$set(root.dir = normalizePath(file.path(here(), 'data')))
# set output directory if saving analysis figures
# odir <- file.path(here(), 'figures')
# dir.create(odir, showWarnings = FALSE)
```


## Simple time series

Let's look at fundamental time series plots, using ggplot2's 'economics' dataset, a data frame with 574 rows and 6 variables:

__date__ - Month of data collection

__pce__ - personal consumption expenditures, in billions of dollars

__pop__ - total population, in thousands

__psavert__ - personal savings rate

__uempmed__ - median duration of unemployment, in weeks

__unemploy__ - number of unemployed in thousands

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
head(economics)
class(economics$pop)
class(economics$date)

base_date_str <- c("1980-06-29")
base_date <- as.Date(base_date_str)

library(lubridate)
lube_date <- ymd(base_date_str)

# Is the outcome of base R and lubridate equivalent?
base_date == lube_date
```

### Plot a single simple time series - what can we do with it?

```{r}
# Basic line plot for num x num time series
ggplot(data = economics, aes(x = date, y = pop))+
  geom_line()
```

```{r}
# adjust the theme to reduce visual background noise
theme_set(theme_minimal())

# play with color and line thickness
ggplot(data = economics, aes(x = date, y = pop))+
  geom_line(color = "#00AFBB", size = 2)
```

```{r}
# adjust the line itself to convey additional diensions of information
ggplot(data = economics, aes(x = date, y = pop)) +
  geom_line(aes(size = unemploy/pop), color = "#FC4E07")
```

```{r}
# Focus on last 10 years only -> Plot a subset of the data
ss <- subset(economics, date > ymd("2006-1-1"))
ggplot(data = ss, aes(x = date, y = pop)) + 
  geom_line(aes(size = unemploy/pop), color = "#FC4E07") +
  labs(size = "unemployed /\npopulation")
```

```{r}
# Focus on post-crash years only -> Plot a subset of the data
ss <- subset(economics, date > ymd("2009-1-1") & date < ymd("2011-1-1"))
ggplot(data = ss, aes(x = date, y = pop)) + 
  geom_line(aes(size = unemploy/pop), color = "#FC4E07") +
  labs(size = "unemployed /\npopulation")
```

```{r}
# Focus on post-crash years only -> Plot a subset of the data
ss <- subset(economics, date > ymd("2009-1-1") & date < ymd("2011-1-1"))
ggplot(data = ss, aes(x = date, y = pop)) + 
  geom_line(aes(color = uempmed, size = unemploy/pop)) +
  labs(color = "Med. weeks\nunemployed", size = "unemployed /\npopulation")
```

## Numerical & Categorical simple series

Let's see some examples of types of series.

### Numerical time, Numerical values - heart rate

- To add colibri, go to https://github.com/bwrc/colibri and follow instructions in the README
- If there is a problem installing devtools in RStudio, try to do it in R. To start R from the command line:
  - Linux, Mac = $ R
  - Windows = find your path to R.exe, e.g. C:\Program Files\R\R-3.6.3\bin>R.exe

```{r, warning=FALSE, message=FALSE}
library(colibri)
```

```{r}
# Add the provided data file to the same folder as the Rmd document
fname <- "DCVRa_D_both_0001-ECG-30s.edf"
## read the data
recording <- read.data.edf(fname)
recording$signal$ECG$data <- -recording$signal$ECG$data  ## need to flip the polarity


## detect the R-peaks
recording <- tryCatch({
  ibi_from_ecg(recording)
}, error = function(e){ 
  print(paste(i, ":: FAIL ::", fname, ":: MYSTERIOUS ERROR IN ibi_from_ecg()!!"))
  return(e)
})
if(inherits(recording, "error")) next


## remove artifacts
ibi_orig <- recording$signal$ibi

ind.artifact <- detect_artifacts_xu(recording$signal$ibi$data)
if (length(ind.artifact) > 0)
{
  recording$signal$ibi$data <- recording$signal$ibi$data[-ind.artifact]
  recording$signal$ibi$t <- recording$signal$ibi$t[-ind.artifact]
}


## plot ecg with R-peaks
par(mfrow = c(3, 1), mai = c(0.3, 1, 0.1, 0.1), mgp = c(1.4, 0.5, 0))
plot(recording$signal$ECG$t, recording$signal$ECG$data, type = "l", col = "blue", lwd = 1
     , xlab = NA, ylab = "uV")
points(recording$signal$ibi.amp$t,recording$signal$ibi.amp$data, type = "p", col = "red", cex = 2, pch = 20)
plot(ibi_orig$t, ibi_orig$data, type = "l", col = "red", lwd = 1, xlab = NA, ylab = "original IBI")
plot(recording$signal$ibi$t, recording$signal$ibi$data, type = "l", col = "green", lwd = 1
     , xlab = "Time (sec)", ylab = "de-artefacted IBI")
```

### Numerical time, Categorical values -  governing parties
```{r}
head(presidential)

ggplot(data = presidential, aes(x = start, y = party)) + 
  geom_point(aes(color = party))
```

```{r}
yrng <- range(economics$unemploy)
ggplot(economics, aes(x = date, y = unemploy)) + 
  geom_line() + 
  geom_rect(aes(NULL, NULL, xmin = start, xmax = end, fill = party),
                   ymin = yrng[1], ymax = yrng[2], data = presidential) + 
  scale_fill_manual(values = alpha(c("blue", "red"), .3))
```

### Categorical time, Numerical values - opening hours
```{r}
alko <- as.data.frame(cbind(matrix(t(rep(9:21, 5)), ncol = 5), c(9:18, rep(NA, 3)), NA))
names(alko) <- c("Ma", "Ti", "Ke", "To", "Pe", "La", "Su")
df <- pivot_longer(alko, 1:7, names_to = "day")
df$day <- factor(df$day,levels = c("Ma", "Ti", "Ke", "To", "Pe", "La", "Su"))

ggplot(data = df, aes(x = day, y = value)) +
  geom_point() +
  ylab("Alko opening hours")
```
### Categorical time, Categorical values

_A special prize will be given to the person who comes up with the best time-series visualisation of Categorical time, Categorical values data!_


## Multiple simple series

Let's see how to plot more than one line per figure:

```{r}
df <- economics %>%
  select(date, psavert, uempmed) %>%
  gather(key = "variable", value = "value", -date)
head(df, 3)
```

```{r}
# Multiple line plot
ggplot(df, aes(x = date, y = value)) + 
  geom_line(aes(color = variable), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  theme_minimal()
```

```{r}
# Area plot
p <- ggplot(df, aes(x = date, y = value)) + 
  geom_area(aes(color = variable, fill = variable), 
            alpha = 0.5, position = position_dodge(0.8)) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  scale_fill_manual(values = c("#00AFBB", "#E7B800")) +
  theme_minimal()
p
```

```{r}
# Format : month/year
p + scale_x_date(date_labels = "%b/%Y")
```

```{r}
# Fitting curves per group
p + stat_smooth(aes(group = variable),
  color = "#FC4E07", fill = "#FC4E07",
  method = "loess"
  )
```


## Characteristics of time series

Let's see how to decompose :

```{r}
# Here's a stationary process
white_noise<-rnorm(100,0,1)
plot(white_noise,type='l',xlab='Time',ylab=expression('x'[t]),main='White Noise Process')
```


```{r}
# Here's a trend-stationary process
white_noise<-rnorm(100,0,40)
trend<-3*(1:100)
x_t<-trend+white_noise
plot(x_t,type='l',xlab='Time',ylab=expression('x'[t]),main='Trend Stationary Process')
```

Assuming that we know that trend stationarity holds, we do a three step process:

1. Fit a model M_t to x_t
2. Subtract M_t from x_t, obtaining y_t = x_t - M_t
3. Model y_t (for inference, forecasting, etc)

```{r}
# Let's detrend and plot
mu_t<-lm(x_t~c(1:100))
y_t<-x_t-mu_t[['fitted.values']]
plot(y_t,type='l',xlab='Time',ylab=expression('Fitted y'[t]),main='Detrended')
```


## Final example

Unique and uniform time series from non-unique, non-uniform generating processes.

```{r}
kings <- scan("http://robjhyndman.com/tsdldata/misc/kings.dat",skip=3)
kings.ts <- ts(kings)
class(kings.ts)
kings.ts
```

```{r}
plot.ts(kings.ts)
```
